Making Sequential Programs Deterministic

### 2019-03-26 design.tex
defaulting and type refinement

@@

@@

\subsection{Discussion on defaulting rules}\label{defaulting}
In our determinism checker, we chose \<@PolyDet> as the default for all method parameters and return types including constructors.
The rationale for this design choice was to not make our type checker too restrictive by making everything \<@Det>.  This implies that
the checker still allows some non-determinism as long as it doesn't break any determinism specifications explicitly written by the programmer.

\begin{itemize}
    \item Example of a method with params and return types annotated as \<@PolyDet>.
    This method called from 2 places - one harmless(@NonDet instantiations but no printing), and another that call with \<@NonDet> args and prints the return type.
    \item Exception to this defaulting is the construction of HashSets and HashMaps because they can't be \<@Det>.
    We explicitly annotate the return type on constructors as
        \<@OrderNonDet> and disallow constructing \<@Det HashSet> or \<@Det
        HashMap>.
\end{itemize}

@@

>>>>

\subsection{Discussion on defaulting rules}\label{defaulting}
Every type system in the checker framework must define a default type annotation. This annotation will be applied
to every location that isn't explicitly annotated. In our determinism checker, we chose \<@PolyDet> as the default for all 
method parameters and return types including constructors.
The rationale for this design choice was to not make our type checker too restrictive by making everything \<@Det>.  This implies that
the checker still allows some non-determinism as long as it doesn't break any determinism specifications explicitly written by the programmer.

\begin{verbatim}
@PolyDet int foo(@PolyDet int x) { }
void bar(@NonDet int random, @Det int value) {
@NonDet int x = foo(random);      // No error.
@Det int y = foo(value);          // No error.
System.out.println(foo(random));  // Error.
}
\end{verbatim}
In the example above, the method \codeid{foo} is annotated with the \<PolyDet> type qualifier which allows
the caller method \codeid{bar} to invoke it with both \<Det> and \<NonDet> arguments. The checker flags an error 
only when the returned value is used in a way that breaks any deterministic specification (Printing a \<NonDet> value
breaks our determinism guarantees).

@@

With the above rules, the checker framework will automatically annotate a local collection with type \<@NonDet> and its
type arguments will get the type \<@Det>. For example, a locally declared \codeid{List} of \codeid{Strings} will be automatically annotated
as \codeid{@NonDet List<@Det String>}. Recall from~\ref{collection-rules} that the determinism checker does not allow this type on
any collection. While this is still sound, it will result in the determinism checker reporting a lot of errors because every locally declared 
collection gets annotated with this invalid type. To avoid this situation, we make an exception to the defaulting rules for local collections
and annotate their type parameters also with \<@NonDet>. So a local list of Strings gets the default annotation of \codeid{@NonDet List<@NonDet String>}. While this defaulting eliminates the possibility of automatically annotating locals collections with an invalid
type, it would could in a high number of false positives. In the following code snippet,
\begin{verbatim}
void test(@Det List<@Det String> argList) {
List<String> localList = argList;
}    
\end{verbatim}
the checker would report an error at the assignment statement since the inferred type for \codeid{localList} is 
\codeid{@NonDet List<@NonDet> String} and since collections types are invariant.
We therefore recommend that programmers explicitly annotate
all local variables that are collections to reduce false positives. The same reasoning and recommendation applies to arrays.

@@

@@

\begin{itemize}
    \item Exception to this defaulting is the construction of HashSets and HashMaps because they can't be \<@Det>.
    We explicitly annotate the return type on constructors as
    \<@OrderNonDet> and disallow constructing \<@Det HashSet> or \<@Det
    HashMap>.
\end{itemize}
